---
title: "Lewisii SAS Cox models and R analysis"
output: Lewis_Flax_Cox_and_Mixed_Modeling
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## SAS Code

## Mixed Modeling of Height

excel file used: **FlaxFreezeForAnalysis.xlsx**

Sheet: **V_DS_Final2**

```{SAS Height, echo=FALSE}
/* Height Data Analysis Start below */
PROC IMPORT OUT= WORK.height  /* SAS dataset name */
            DATAFILE= "~\FlaxFreezeForAnalysis.xlsx" 
            DBMS=XLSX REPLACE;
     SHEET="V_DS_Final2"; /* input dataset tab in excel sheet */
     GETNAMES=YES;
RUN;
quit;
ods trace off;
ods select all/* exclude IntPlot (persist) BoxPlot (persist)*/;
ods listing close;
ods html path='C:\Users\Documents\YourFolderPathHere\' file='HeightAnalysisAR.xls' style=default;

ods html file="Flax Height resid Analysis.htm" path="C

proc sort data=height;
by Sample_Number rep treat Time;
run;
proc mixed cl covtest data=height PLOTS=ResidualPanel(UNPACKBOX);
class rep treat Time;
model Total_HeightRegBrady = treat | Time /outp=resid2 ddfm=satterth; /* the | symbol will model model main effects and interactions (useful for 3way+ interactions.  outp tests residuals (ddfm determines degrees of freedom*/
random rep;
repeated Time/subject=rep*treat type=ar(1) rcorr; /* un ins unstuctured as the correlations between and amoung weeks may have different covariance?' */
by Sample_Number;
run;

/* Start box cox and shapiro wilks and boxcox below ##############################*/*//;
proc sort data=resid2;
by Sample_Number;
run;


proc univariate data=resid2 normal;
var resid;
by Sample_Number;
run;
quit;


/* Heritability For 2WP analysis and DOT  */ 
/* Not used in published article */
PROC IMPORT OUT= WORK.height  /* SAS dataset name */
            DATAFILE= "~C:\FlaxFreezeForAnalysis.xlsx" /* input data set filepath */
            DBMS=XLSX REPLACE;
     SHEET="Sheet1"; /* input dataset tab in excel sheet */
     GETNAMES=YES;
RUN;
quit;
proc sort data=height;
by Accession rep ;
run;
proc mixed data =height;
class rep Accession;
model HeightDOT =  ;
random rep Accession;
run;
quit;
*/ End Heritability analysis /*


### Testing Residuals and compare Tests

ods html file="Flax Height TRANSresidAnalysis.htm" path="C:\Users\Documents\FixedResidualsFilePath" gpath="C:\Users\Documents\FixedResidualsFilePath";
ods html path='C:\Users\Documents\FixedResidualsFilePath' file='HeightAnalysisARResidAdj.xls' style=default;
proc transreg data=height plots=box;
model BoxCox (Total_HeightRegBrady) = class(rep treat treat*Time Time);
by Sample_Number;
/*output out=height2;*/
run; 

%macro datatransform (sample, lambda);
data height;
set height;
if Sample_Number = &sample and &lambda ^= 0 then t_height=((Total_HeightRegBrady**(&lambda))-1)/&lambda;
run;
%mend;
%datatransform(1, 0.25); 
%datatransform(2,1);
%datatransform(3,1);
%datatransform(4,1.5);
%datatransform(5,0.75);
%datatransform(6,1.5);
%datatransform(7,0);
%datatransform(8,1);
%datatransform(9,1.25);
%datatransform(10,1);
%datatransform(11,0);
%datatransform(12,1.75);
%datatransform(13,0.5);
%datatransform(14,0.75)
%datatransform(15,0.75);
%datatransform(16,0.25);
%datatransform(17,0.75);
%datatransform(18,0.75);
%datatransform(19,1.25);
%datatransform(20,1);
%datatransform(21,0.75);
%datatransform(22,0.5);
%datatransform(23,0.75);
%datatransform(24,0.5);
%datatransform(25,0.5);
%datatransform(26,1);
%datatransform(27,0.75);
%datatransform(28,0.25);
%datatransform(29,1);
%datatransform(30,1.75);
%datatransform(31,1.5);
%datatransform(32,-0.5);
%datatransform(33,2.25);
%datatransform(34,0.75);
%datatransform(35,2);
%datatransform(36,0.75);
%datatransform(37,0.75);
%datatransform(38,0.5);
%datatransform(39,0);
%datatransform(40,1.25);
%datatransform(41,0.25);
%datatransform(42,0.25);
%datatransform(43,0.25);
%datatransform(44,2);
%datatransform(45,1.25);
%datatransform(46,1);

proc mixed cl covtest data=height PLOTS=ResidualPanel(UNPACKBOX);
class rep treat Time;
model t_height = treat | Time /outp=resid3 ddfm=satterth; /* the | symbol will model model main effects and interactions (useful for 3way+ interactions.  outp tests residuals (ddfm determines degrees of freedom*/
random rep;
repeated Time/subject=rep*treat type=ar(1) rcorr; /* un ins unstuctured as the correlations between and amoung weeks may have different covariance?' */
by Sample_Number;
run;
proc univariate data=resid3 normal plot;
var resid;
by Sample_Number;
probplot resid /normal noframe;
run;

quit;
```

## Cox Proportional Hazards

Data file needed: FlaxFreezeForAnalysis.xlsx or trialcoxNO.csv

Sheet: V_DS_Final

```{SAS}

data work.cox;
infile "/Users\Fleur\Documents\Thesis Files\Freeze Study\trialcoxNO.csv"
delimiter = ","
missover
dsd
firstobs = 2;

format Cell_Number best12.;
format Sample_Number best12.;
format treat best12.;
format rep best12.;
format SurvivalRegulated best12.;
format Time best12.;

input
Cell_Number $
Sample_Number $
treat $
rep $
SurvivalRegulated $
Time $;

run;

data cox;
set cox;
if Sample_Number="20" then delete;
if Sample_Number="26" then delete;
if Sample_Number="28" then delete;
if Sample_Number="46" then delete;
run;

/*alternative data input Below ######################## 20 26 28 46 already removed from V_DS_Final  ########################*/

PROC IMPORT OUT= WORK.cox  /* SAS dataset name */
            DATAFILE= "C:\Users\Fleur\Documents\Thesis Files\Freeze Study\FlaxFreezeForAnalysis.xlsx" /* input data set filepath */
            DBMS=XLSX REPLACE;
     SHEET="V_DS_Final"; /* input dataset tab in excel sheet */
     GETNAMES=YES;
RUN;
quit;

data cox;
set cox (drop=rep Freeze_Damage);
run;
proc sort data=cox;
by Time;
run;
data cox2;
set cox (drop=treat);
run;

ods graphics on;
ods excel file="C:\Users\Documents\YourCoxSummaryFileNameAndPathHERE.xlsx"; /* html path="C:\Users\Documents\" file="YourCoxSummaryFileNameAndPathHERE.xlsx" style=default; */

proc phreg data=cox plots(overlay)=survival;
class Sample_Number(ref='45');
model treat*SurvivalRegulated(1)=Sample_Number / TIES=Efron;
/*id Sample_Number treat;*/
/* id statement specifies additional variables for identifying observations in the input data and are placed in the out= dataset */
baseline covariates=cox2 out=Estimate3 CUMHAZ=_ALL_ survival=_all_/ rowid=Sample_Number;
hazardratio Sample_Number / diff=DISTINCT cl=BOTH;
by Time;
run;

proc sort data=Estimate3;
by Sample_Number treat Survival Time;
run;
ods excel close;
quit;

/*pull unique rows to eliminate the repeated data from the generated table. */
data first_unique_rows;
set Estimate3;
by Sample_Number treat Survival Time;
if first.Survival and first.Time; run; 

proc sort data=first_unique_rows;
by Sample_Number Time;
run;

data first_unique_rows;
set first_unique_rows (drop=Total_Dead_Stems Total_Stems Total_HeightRegBrady);
run;

```

## C**orrelation Data Transformations**

Data transformation performed by SAS and R for height data for correlation tests is listed below. **Orders of operations may not be conserved** as alternation between the two software was done multiple times. It is **not recommended** to perform these transformations this way, instead I recommend to remain in one software for all data steps. The steps below were provided for transparency only.

```{SAS}
PROC IMPORT OUT= WORK.HeightAvg  /* SAS dataset name */
            DATAFILE= "C:\Users\Fleur\Documents\Thesis Files\Freeze Study\HeightAvg.xlsx" /* input data set filepath */
            DBMS=XLSX REPLACE;
     SHEET="Sheet1"; /* input dataset tab in excel sheet */
     GETNAMES=YES;
RUN;
quit;
proc sort data = Heightavg;
by Sample_Number treat Time;
run;
proc transpose data = Heightavg out=Heightavg_wide(drop=_name_ _LABEL_);
by Sample_Number treat;
var Total_HeightRegBrady;
id Time;
run;
proc export data=Heightavg_wide
outfile="C:\Users\Fleur\Documents\Thesis Files\Freeze Study\Heightavg_wide.csv" /* input data set filepath */
            DBMS=csv 
replace; run;

/* Start of Correlation testing ######################################## */

PROC IMPORT OUT= WORK.VDS  /* SAS dataset name */
            DATAFILE= "C:\Users\Fleur\Documents\Thesis Files\Freeze Study\Correlation tests\FlaxGeoInfo.xlsx " /* input data set filepath */
            DBMS=XLSX REPLACE;
     SHEET="V_DS_Final"; /* input dataset tab in excel sheet */
     GETNAMES=YES;
RUN;
quit; 

proc sort data = VDS;
by Sample_Number treat Time;
run;
proc transpose data = VDS out=VDS_wide(drop=_name_ _LABEL_);
by Sample_Number treat rep;
var SurvivalRegulated;
id Time;
run;

proc export data=VDS_wide
outfile="C:\Users\Fleur\Documents\Thesis Files\Freeze Study\Correlation tests\VDS_wide.csv" /* input data set filepath */
            DBMS=csv 
replace; run;

PROC IMPORT OUT= WORK.SurvAvg  /* SAS dataset name */
            DATAFILE= "C:\Users\Fleur\Documents\Thesis Files\Freeze Study\Correlation tests\SurvTry2.xlsx " /* input data set filepath */
            DBMS=XLSX REPLACE;
     SHEET="Sheet 1"; /* input dataset tab in excel sheet */
     GETNAMES=YES;
RUN;
quit; 

proc sort data=SurvAvg ;
by Sample_Number treat Time;
run;
proc transpose data = SurvAvg out=SurvAvg_wide (drop=_name_ _LABEL_) ;
by Sample_Number treat;
var SurvivalRegulated;
id Time;
run;
proc export data=SurvAvg_wide
outfile="C:\Users\Fleur\Documents\Thesis Files\Freeze Study\Correlation tests\SurvAvg_wide.csv" /* input data set filepath */
            DBMS=csv 
replace; run;


/* Export to R to get averages easily then reimport to run corr. */

proc export data=Vds_wide
outfile="C:\Users\Fleur\Documents\Thesis Files\Freeze Study\Correlation tests\Surv_wide.csv" /* input data set filepath */
            DBMS=csv 
replace; run;

PROC IMPORT OUT= WORK.HeightDiff  /* SAS dataset name */
            DATAFILE= "C:\Users\Fleur\Documents\Thesis Files\Freeze Study\Correlation tests\FlaxGeoInfo.xlsx " /* input data set filepath */
            DBMS=XLSX REPLACE;
     SHEET="V_DS_Final"; /* input dataset tab in excel sheet */
     GETNAMES=YES;
RUN;
quit; 

proc sort data=HeightDiff ;
by Sample_Number rep treat Time;
run;
proc transpose data = HeightDiff out=HeightDiff_wide (drop=_name_ _LABEL_) ;
by Sample_Number rep treat;
var Total_HeightRegBrady;
id Time;
run;

data HeightDiff2_wide;
set HeightDiff_wide;
BaselineOne = (_1-_0);
BaselineTwo = (_2-_1);
BaselineThree = (_3-_2);
run;
quit;

proc export data=HeightDiff2_wide
outfile="C:\Users\Fleur\Documents\Thesis Files\Freeze Study\Correlation tests\HeightDifAvg_wide.csv" /* input data set filepath */
            DBMS=csv 
replace; run;


```

The R code for height and survival means

new libraries used: library(openxlsx), library(writexl)

```{r}
library(dplyr)
setwd("C:/Users/Fleur/Documents/Thesis Files/Freeze Study/Correlation tests")
WideSurvData=read.csv("Surv_wide.csv", sep = ",")
SURVTIME<-WideSurvData
head(SURVTIME)
colnames(SURVTIME)<- c('Sample_Number','treat','0','1','2','3')
WideSurvData %>% group_by(Sample_Number, treat) %>% 
  summarise_at(vars("ST3"),mean, na.rm=T) -> SURVTIME3


WideTry2=read.csv("VDSExcel.csv", sep = ",")
#Vertical_MDS in FlaxFreezeforAnalysis
WideTry2 %>% group_by(Sample_Number, Time, treat) %>% 
  summarise_at(vars("SurvivalRegulated"),mean, na.rm=T) -> SurvTry2
WideTry2 %>% group_by(Sample_Number, Time, treat) %>% 
  summarise_at(vars("Freeze_Damage"),mean, na.rm=T) -> FreezeAvg
HeightWide=read.csv("HeightDifAvg_wide.csv", sep = ",")
HeightWide %>% group_by(Sample_Number, treat) %>% 
  summarise_at(vars("BaselineOne"),mean, na.rm=T) ->Base1Avg
HeightWide %>% group_by(Sample_Number, treat) %>% 
  summarise_at(vars("BaselineTwo"),mean, na.rm=T) ->Base2Avg
HeightWide %>% group_by(Sample_Number, treat) %>% 
  summarise_at(vars("BaselineThree"),mean, na.rm=T) ->Base3Avg

Column2toadd<- Base2Avg$BaselineTwo
Column3toadd<- Base3Avg$BaselineThree
Base1Avg<-cbind(Base1Avg,Column2toadd,Column3toadd)
colnames(Base1Avg)<- c('Sample_Number','treat','0to1Avg','1to2Avg','2to3Avg')

library(openxlsx)
openxlsx::write.xlsx(SurvTry2, "SurvTry2.xlsx", rowNames=FALSE)
openxlsx::write.xlsx(FreezeAvg, "FreezeAvg.xlsx", rowNames=FALSE)
openxlsx::write.xlsx(Base1Avg, "HeightGrowthAvg.xlsx", rowNames=FALSE)


```

## C**orrelation Test Data Transformations**

File needed: **FlaxGeoInfo.xlsx**

Sheets: FullWideAvg, 9only, 12only, 15only

```{SAS Correlations}
PROC IMPORT OUT= WORK.cortest  /* SAS dataset name */
            DATAFILE= "C:\Users\Fleur\Documents\Thesis Files\Freeze Study\Correlation tests\FlaxGeoInfo.xlsx " /* input data set filepath */
            DBMS=XLSX REPLACE;
     SHEET="FullWideAvg"; /* input dataset tab in excel sheet */
     GETNAMES=YES;
RUN;
quit; 


Proc corr data=cortest pearson plots=scatter(nvar=all); 
 Var ElevationM Latitude Longitude Hazard _0to1Avg _1to2Avg _2to3Avg HT_0 HT_1 HT_2 HT_3 Freeze_Damage ;
run;
ods output Correlations=cortest;
quit;

/* Correlations by treatments below */


PROC IMPORT OUT= WORK.only9corr  /* SAS dataset name */
            DATAFILE= "C:\Users\Fleur\Documents\Thesis Files\Freeze Study\Correlation tests\FlaxGeoInfo.xlsx " /* input data set filepath */
            DBMS=XLSX REPLACE;
     SHEET="9only"; /* input dataset tab in excel sheet */
     GETNAMES=YES;
RUN;
quit; 
Proc corr data=only9corr pearson plots=scatter(nvar=all); 
 Var ElevationM Latitude Longitude Hazard _0to1Avg _1to2Avg _2to3Avg HT_0 HT_1 HT_2 HT_3 ;
run;
ods output Correlations=only9corr;

PROC IMPORT OUT= WORK.only15corr  /* SAS dataset name */
            DATAFILE= "C:\Users\Fleur\Documents\Thesis Files\Freeze Study\Correlation tests\FlaxGeoInfo.xlsx " /* input data set filepath */
            DBMS=XLSX REPLACE;
     SHEET="15only"; /* input dataset tab in excel sheet */
     GETNAMES=YES;
RUN;
quit; 
Proc corr data=only15corr pearson plots=scatter(nvar=all); 
 Var ElevationM Latitude Longitude Hazard _0to1Avg _1to2Avg _2to3Avg HT_0 HT_1 HT_2 HT_3 ;
run;
ods output Correlations=only15corr;
PROC IMPORT OUT= WORK.only12corr  /* SAS dataset name */
            DATAFILE= "C:\Users\Fleur\Documents\Thesis Files\Freeze Study\Correlation tests\FlaxGeoInfo.xlsx " /* input data set filepath */
            DBMS=XLSX REPLACE;
     SHEET="12only"; /* input dataset tab in excel sheet */
     GETNAMES=YES;
RUN;
quit; 
Proc corr data=only12corr pearson plots=scatter(nvar=all); 
 Var ElevationM Latitude Longitude Hazard _0to1Avg _1to2Avg _2to3Avg HT_0 HT_1 HT_2 HT_3 ;
run;
ods output Correlations=only12corr;
```
